### 2016-05-29 text/Chapter01.tex
Spell checking
Addressing feedback, fix demands (Chapter 6)

To support cross-compilation, the language needs to be context-aware. Each of the function has a
number of context requirements. The \ident{fetchNews} function needs to have access to a database;
\ident{fetchLocalNews} needs access to a GPS sensor and to a network (to perform the remote call).
However, it does not need a specific platform -- it can work on both iPhone and Windows. The last
two platform-specific functions inherit the requirements of \ident{fetchLocalNews} and additionally
also require a specific platform.

@@

The approach advocated in this thesis is to track information about context requirements using
the type system. To make this practical, the system should also provide at least a partial support for
automatic type inference, as the information about context requirements makes the types more complex.
An inspiring example might be the F\# support for units of measure \cite{app-units} -- the user has
to explicitly annotate constants, but the rest of the information is inferred automatically.

@@

\noindent
Here, we use the notation $\tau_1 \,\ident{\hspace{-1em}\footnotesize{@}}\, c \rightarrow \tau_2$ to denote
a function that takes an input of type $\tau_1$, produces a result of type $\tau_2$ and has additional
context requirements specified by $c$. In the above example, the annotation $c$ is simply a set of
required resources or capabilities. However, a more complex structure could be used as well, for example,
including the Android API level as an integer.

@@

\paragraph{Tracking context requirements.}

The systems for tracking of context requirements developed in this thesis are inspired by the idea
of effect systems. To demonstrate our approach, consider the following call from the sample program
shown earlier -- first using standard ML-like type system:
%
\begin{equation*}
\ident{password}\!:\!\ident{string},~\ident{cmd}\!:\!\ident{string}
  \vdash \ident{query}(\ident{cmd}, \ident{password}) : \ident{news list}
\end{equation*}
%
The expression queries a database and gets back a list of news values as the result. Recall from the
earlier discussion that there are two contextual information that are desirable to track for this
expression. First, the call to the \ident{query} primitive requires \emph{database access}. Second,
the \ident{password} argument needs to be marked as \emph{sensitive value} to avoid sending it over
an unsecure network connection. The \emph{coeffect systems} developed in this thesis capture this
information in the following way (we slightly refine the notation later):
%
\begin{equation*}
\begin{array}{l}
 \coctx{( \ident{password}\!:\!\coctx{\ident{string}}{\hspace{-0.75em}\ident{\cclrd{\footnotesize sensitive}}},
   \ident{cmd}\!:\!\ident{string})}{ \aclrd{\{ \ident{database} \}}}  \\[-0.1em]
 \hspace{3em} \vdash\ident{query}(\ident{cmd}, \ident{password}) : \ident{news list}
\end{array}
\end{equation*}
%
Rather than attaching the annotation to the \emph{resulting type}, we attach them to the variable
context $\Gamma$. In other words, coeffect systems do not keep track just of the variables available
in the context -- they also capture detailed information about the execution environment.
In the above example, the system tracks metadata about the variables and annotates \ident{password}
as sensitive. Furthermore, it tracks requirements about the execution environment, for example,
that the execution requires an access to database.

@@

\item Related work is presented in Chapter~\ref{ch:pathways} and, together with further work,
  throughout the thesis. Two important directions deserve furhter exploration. In Chapter~\ref{ch:further},
  we outline a unified coeffect system that is capable of capturing both flat and structural
  properties. We also include a brief discussion of a different approach to tracking contextual
  information that arises from modal logics.
\end{itemize}

>>>>

To support cross-compilation, the language needs to be context-aware. Each of the function has a
number of context demands. The \ident{fetchNews} function needs to have access to a database;
\ident{fetchLocalNews} needs access to a GPS sensor and to a network (to perform the remote call).
However, it does not need a specific platform -- it can work on both iPhone and Windows. The last
two platform-specific functions inherit the requirements of \ident{fetchLocalNews} and additionally
also require a specific platform.

@@

The approach advocated in this thesis is to track information about context demands using
the type system. To make this practical, the system should also provide at least a partial support for
automatic type inference, as the information about context demands makes the types more complex.
An inspiring example might be the F\# support for units of measure \cite{app-units} -- the user has
to explicitly annotate constants, but the rest of the information is inferred automatically.

@@

\noindent
Here, we use the notation $\tau_1 \,\ident{\hspace{-1em}\footnotesize{@}}\, c \rightarrow \tau_2$ to denote
a function that takes an input of type $\tau_1$, produces a result of type $\tau_2$ and has additional
context demands specified by $c$. In the above example, the annotation $c$ is simply a set of
required resources or capabilities. However, a more complex structure could be used as well, for example,
including the Android API level as an integer.

@@

\paragraph{Tracking context demands.}

The systems for tracking of context demands developed in this thesis are inspired by the idea
of effect systems. To demonstrate our approach, consider the following call from the sample program
shown earlier -- first using standard ML-like type system:
%
\begin{equation*}
\ident{password}\!:\!\ident{string},~\ident{cmd}\!:\!\ident{string}
  \vdash \ident{query}(\ident{cmd}, \ident{password}) : \ident{news list}
\end{equation*}
%
The expression queries a database and gets back a list of news values as the result. Recall from the
earlier discussion that there are two contextual information that are desirable to track for this
expression. First, the call to the \ident{query} primitive requires \emph{database access}. Second,
the \ident{password} argument needs to be marked as \emph{sensitive value} to avoid sending it over
an unsecure network connection. The \emph{coeffect systems} developed in this thesis capture this
information in the following way (we slightly refine the notation later):
%
\begin{equation*}
\begin{array}{l}
 \coctx{( \ident{password}\!:\!\coctx{\ident{string}}{\hspace{-0.75em}\ident{\cclrd{\footnotesize sensitive}}},
   \ident{cmd}\!:\!\ident{string})}{ \aclrd{\{ \ident{database} \}}}  \\[-0.1em]
 \hspace{3em} \vdash\ident{query}(\ident{cmd}, \ident{password}) : \ident{news list}
\end{array}
\end{equation*}
%
Rather than attaching the annotation to the \emph{resulting type}, we attach them to the variable
context $\Gamma$. In other words, coeffect systems do not keep track just of the variables available
in the context -- they also capture detailed information about the execution environment.
In the above example, the system tracks metadata about the variables and annotates \ident{password}
as sensitive. Furthermore, it tracks requirements about the execution environment, for example,
that the execution requires an access to database.

@@

\item Related work is presented in Chapter~\ref{ch:pathways} and, together with further work,
  throughout the thesis. Two important directions deserve further exploration. In Chapter~\ref{ch:further},
  we outline a unified coeffect system that is capable of capturing both flat and structural
  properties. We also include a brief discussion of a different approach to tracking contextual
  information that arises from modal logics.
\end{itemize}

### 2016-05-29 text/Chapter02.tex
Addressing feedback, fix demands (Chapter 6)

\noindent
When we view variable access as a context requirement, we can see that the body of the function
($\ident{x}+\ident{y}$) requires a context that provides values for variables \ident{x} and
\ident{y}. In static binding, the context requirements of the body can be placed on the scope in
which the function is defined (declaration-site). In dynamic binding, the requirements are
\emph{delayed} and are placed on the scope in which the function is called (call site).

@@

\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{f}=\\[-0.25em]
\qquad\kvd{let}~\ident{?x}=10~\kvd{in}\\[-0.25em]
\qquad\lambda \ident{y}\rightarrow \ident{?x}+\ident{y}~\kvd{in}\\[-0.25em]
\kvd{let}~\ident{?x}=5~\kvd{in}\\
\ident{f}~0
\end{array}\hspace{5em}\begin{array}{l}
\kvd{let}~\ident{f}=\\[-0.25em]
\qquad\lambda \ident{y}\rightarrow \ident{?x}+\ident{y}~\kvd{in}\\[-0.25em]
\kvd{let}~\ident{?x}=5~\kvd{in}\\
\ident{f}~0
\end{array}
\end{equation*}
%
The binding rules for implicit parameters can be summarized as ``\emph{static binding when possible,
dynamic binding when needed}''. If an implicit parameter is available in scope, then the value
is statically bound and the context requirement is satisfied using the declaration-site context.
Otherwise, the context requirement is delayed and has to be satisfied at the call site.
In the example on the left, \ident{?x} is bound to $10$ the function \ident{f} has no
delayed context requirements and the expression evaluates to $10$. In the example on the right,
the context requirement \ident{?x} is delayed and is satisfied via dynamic binding when calling
the function. The expression evaluates to $5$.

In Haskell, the type system checks that bindings for all required implicit parameters are
available. The type of the function \ident{f} on the left is $\kvd{int}\rightarrow\kvd{int}$, while
the type of the \ident{f} function on the right is $\{\ident{?x}:\kvd{int}\}\Rightarrow\kvd{int}\rightarrow\kvd{int}$.
The part before $\Rightarrow$ specifies the required implicit parameters that need to be available
in the environment when calling the function. It is worth noting that the syntax is similar to the
one used by type-class constraints. Those can be viewed as context requirements too
(Section~\ref{sec:applications-flat-impl}).

\paragraph{Thesis perspective.}
The three different binding mechanisms discussed so far can be seen as different ways of splitting
context requirements of a particular kind. Dynamic and static binding represent the opposite ends
of the design space and Haskell's implicit parameters are an interesting point inside the wider space.

In this thesis, we consider different notions of context, using implicit parameters as just one
of several motivating examples. However, implicit parameters are a valuable example, because they
clearly illustrate the ambiguity inherent in context-aware programs -- the context requirements of
a function can be satisfied using the context available at declaration-site or using the context
available at the call site.

We also aim to find a description of context-aware languages that does not make ad-hoc decisions
about how context requirements are split between the declaration-site and the call site. While
Haskell's solution for implicit parameters might be the most reasonable one, different notions
of context might require different domain-specific choices and the general framework of
context-aware programming should make that possible.

@@

This approach decouples two important aspects of context-aware programming and lets us study them
independently -- the semantics of context-aware programs and the domain-specific way of resolving
how context requirements are satisfied. In our treatment of implicit parameters, we consider
multiple typing derivations (representing a range with static and dynamic scoping at opposite ends),
but we uniquely choose one preferred typing (in case of implicit parameters, mimicking the
behaviour of Haskell).

@@

The language in Figure~\ref{fig:path-coeff} extends simple lambda calculus with resources and
with a construct $\kvd{access}~e$ that obtains the resource specified by the expression $e$.
Most of the typing rules correspond to those of effect systems. Variable access (\emph{var})
has no context requirements, application (\emph{app}) combines context requirements of the two
sub-expressions and latent context-requirements of the function.
The (\emph{abs}) rule is different than the corresponding rule for effect systems -- the
resource requirements of the body $\cclrd{r} \cup \cclrd{s}$ are split between the \emph{immediate
context-requirements} associated with the current context $\coctx{\Gamma}{\cclrd{r}}$ and the
\emph{latent context-requirements} of the function.

@@

\paragraph{Thesis perspective.}
As mentioned earlier, we are interested in designing context-dependent languages and so we
use comonads for \emph{language semantics}. Uustalu and Vene present a semantics of
context-dependent computations in terms of comonads. We provide the rest of the story known
from the marriage of monads and effects. We develop coeffect calculus with an type system that
tracks the context requirements more precisely (by annotating the types) and we add indexing
to comonads and link the two by giving a formal semantics. The indexing allows us to capture
applications that do not fit into the model provided by plain comonads.

>>>>

\noindent
When we view variable access as a context requirement, we can see that the body of the function
($\ident{x}+\ident{y}$) requires a context that provides values for variables \ident{x} and
\ident{y}. In static binding, the context demands of the body can be placed on the scope in
which the function is defined (declaration-site). In dynamic binding, the requirements are
\emph{delayed} and are placed on the scope in which the function is called (call site).

@@

\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{f}=\\[-0.25em]
\qquad\kvd{let}~\ident{?x}=10~\kvd{in}\\[-0.25em]
\qquad\lambda \ident{y}\rightarrow \ident{?x}+\ident{y}~\kvd{in}\\[-0.25em]
\kvd{let}~\ident{?x}=5~\kvd{in}\\
\ident{f}~0
\end{array}\hspace{5em}\begin{array}{l}
\kvd{let}~\ident{f}=\\[-0.25em]
\qquad\lambda \ident{y}\rightarrow \ident{?x}+\ident{y}~\kvd{in}\\[-0.25em]
\kvd{let}~\ident{?x}=5~\kvd{in}\\
\ident{f}~0
\end{array}
\end{equation*}
%
The binding rules for implicit parameters can be summarized as ``\emph{static binding when possible,
dynamic binding when needed}''. If an implicit parameter is available in scope, then the value
is statically bound and the context requirement is satisfied using the declaration-site context.
Otherwise, the context requirement is delayed and has to be satisfied at the call site.
In the example on the left, \ident{?x} is bound to $10$ the function \ident{f} has no
delayed context demands and the expression evaluates to $10$. In the example on the right,
the context requirement \ident{?x} is delayed and is satisfied via dynamic binding when calling
the function. The expression evaluates to $5$.

In Haskell, the type system checks that bindings for all required implicit parameters are
available. The type of the function \ident{f} on the left is $\kvd{int}\rightarrow\kvd{int}$, while
the type of the \ident{f} function on the right is $\{\ident{?x}:\kvd{int}\}\Rightarrow\kvd{int}\rightarrow\kvd{int}$.
The part before $\Rightarrow$ specifies the required implicit parameters that need to be available
in the environment when calling the function. It is worth noting that the syntax is similar to the
one used by type-class constraints. Those can be viewed as context demands too
(Section~\ref{sec:applications-flat-impl}).

\paragraph{Thesis perspective.}
The three different binding mechanisms discussed so far can be seen as different ways of splitting
context demands of a particular kind. Dynamic and static binding represent the opposite ends
of the design space and Haskell's implicit parameters are an interesting point inside the wider space.

In this thesis, we consider different notions of context, using implicit parameters as just one
of several motivating examples. However, implicit parameters are a valuable example, because they
clearly illustrate the ambiguity inherent in context-aware programs -- the context demands of
a function can be satisfied using the context available at declaration-site or using the context
available at the call site.

We also aim to find a description of context-aware languages that does not make ad-hoc decisions
about how context demands are split between the declaration-site and the call site. While
Haskell's solution for implicit parameters might be the most reasonable one, different notions
of context might require different domain-specific choices and the general framework of
context-aware programming should make that possible.

@@

This approach decouples two important aspects of context-aware programming and lets us study them
independently -- the semantics of context-aware programs and the domain-specific way of resolving
how context demands are satisfied. In our treatment of implicit parameters, we consider
multiple typing derivations (representing a range with static and dynamic scoping at opposite ends),
but we uniquely choose one preferred typing (in case of implicit parameters, mimicking the
behaviour of Haskell).

@@

The language in Figure~\ref{fig:path-coeff} extends simple lambda calculus with resources and
with a construct $\kvd{access}~e$ that obtains the resource specified by the expression $e$.
Most of the typing rules correspond to those of effect systems. Variable access (\emph{var})
has no context demands, application (\emph{app}) combines context demands of the two
sub-expressions and latent context-requirements of the function.
The (\emph{abs}) rule is different than the corresponding rule for effect systems -- the
resource requirements of the body $\cclrd{r} \cup \cclrd{s}$ are split between the \emph{immediate
context-requirements} associated with the current context $\coctx{\Gamma}{\cclrd{r}}$ and the
\emph{latent context-requirements} of the function.

@@

\paragraph{Thesis perspective.}
As mentioned earlier, we are interested in designing context-dependent languages and so we
use comonads for \emph{language semantics}. Uustalu and Vene present a semantics of
context-dependent computations in terms of comonads. We provide the rest of the story known
from the marriage of monads and effects. We develop coeffect calculus with an type system that
tracks the context demands more precisely (by annotating the types) and we add indexing
to comonads and link the two by giving a formal semantics. The indexing allows us to capture
applications that do not fit into the model provided by plain comonads.

### 2016-05-29 text/Chapter03.tex
Addressing feedback, fix demands (Chapter 6)

Effect systems generally use judgements of the form $\Gamma \vdash e : \tau \;\&\; \cclrd{\sigma}$,
associating effects $\sigma$ with the output type. We write coeffect
systems using judgements of the form $\coctx{\Gamma}{\cclrd{\sigma}} \vdash e : \tau$, associating
the context requirements with $\Gamma$. Thus, we extend the traditional notion of free-variable
context $\Gamma$ with richer notions of context. This notation emphasizes the right intuition,
but there are more important differences between effects and coeffects.

@@

In Section~\ref{sec:intro-context-example}, we used \emph{resources} in a distributed system as an
example of flat coeffects. These could be, for example, a database, GPS sensor or access to the current
time. We also outlined that such context requirements can be tracked as part of the typing assumption,
for example, say we have an expression $e$ that requires GPS coordinates and the current time.
The variable context of such expression will be annotated with a set of required resources,
\ie~$\coctx{\Gamma}{ \cclrd{\ident{\{ gps, time \}}} }$.

@@

We look at resources in distributed programming in more details in Section~\ref{sec:applications-flat-distr}.
The important point here is that in flat coeffect systems, contextual requirements are
\emph{split} between the call site and declaration site. Furthermore, there is no syntactic
structure that determines how the requirements are split. As mentioned in Section~\ref{sec:path-binding-amb},
we decouple the definition of semantics from the domain-specific choice that determines how
context requirements are satisfied. We capture the choice in the type and give semantics over a
\emph{typing derivation}. A domain-specific algorithm then chooses the desirable typing -- for
example, by preferring resources available on the client over resources available on the server.

@@

Just as in the usual pointwise multiplication of a vector by a scalar, we lift any binary operation on scalars into a
scalar-vector one. For a binary operation on scalars $\circ : \C \times \C \rightarrow \C$, we define
 $\cclrd{s} \circ \aclrd{\textbf{r}} = \alift{ \cclrd{s}\circ\cclrd{r_1}, \ldots, \cclrd{s}\circ\cclrd{r_n}}$.
Relations on scalars can be also lifted to vectors. Given two vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}$ of the
same shape with positions $\{ 1, \ldots, n \}$ and a relation $\varpropto\, \subseteq \C \times \C$ we define
$\aclrd{\textbf{r}} \varpropto \aclrd{\textbf{s}} \Leftrightarrow (\cclrd{r_1} \varpropto \cclrd{s_1}) \wedge \ldots \wedge (\cclrd{r_n} \varpropto \cclrd{s_n}) $
Finally, we often concatenate vectors, for example, when joining two variable contexts.
Given vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}$ with (possibly different) shapes $\{ 1, \ldots, n \}$ and
$\{ 1, \ldots, m \}$, the associative operation for concatenation $\times$ is defined as
$\aclrd{\textbf{r}}\times\aclrd{\textbf{s}} = \alift{\cclrd{r_1},\ldots,\cclrd{r_n},\cclrd{s_1},\ldots,\cclrd{s_m}}$.

We note that an environment $\Gamma$ containing $n$ uniquely named, typed variables is also a vector,
but we continue to write `$,$' for the product, so $\Gamma_1, x\!:\!\tau, \Gamma_2$ should
be seen as $\Gamma_1 \times \langle x\!:\!\tau\rangle \times \Gamma_2$.

@@

Accessing an ordinary variable (\emph{var}) does not require any implicit parameters. The rule that
introduces primitive context requirements is (\emph{param}). Accessing a parameter \ident{?param}
requires it to be available in the context. The context may provide more (unused)
implicit parameters thanks to the subcoeffecting rule (\emph{sub}).

When we read the rules from the top to the bottom, application (\emph{app}) and let binding
(\emph{let}) simply union the context requirements of the sub-expressions. However, lambda abstraction
(\emph{abs}) is where the example differs from effect systems. The implicit parameters required by
the body $\cclrd{r} \cup \cclrd{s}$ can be freely split between the declaration site ($\coctx{\Gamma}{\cclrd{r}}$)
and the call site ($\tau_1 \xrightarrow{\cclrd{s}} \tau_2$) and thus an expression may have multiple
valid typing derivations. Finally, (\emph{letdyn}) removes the bound parameter from the set of requirements.

@@

Type classes and implicit parameters show two important points about flat coeffect systems.
First, the context requirements are associated with some \emph{scope}, such as the body
of a function. Second, they are associated with the input. To call a function that takes an
implicit parameter or has a type-class constraint, the caller needs to pass a (hidden) parameter
together with the function inputs.

@@

\paragraph{Summary.}
Unlike with implicit parameters, lambda abstraction for liveness analysis does not introduce
non-determinism. It simply duplicates the context requirements. However, this still matches the
property of coeffects that impurities cannot be delayed or thunked and attached just to the
function arrow -- we place requirements on both call site and declaration site.

@@

\begin{figure}[t]
a.) Ordinary, syntax-driven rules along with subcoeffecting
\begin{equation*}
\tyrule{var}{}
  {\coctx{x \!:\! \tau}{\alift{\cclrd{1}}} \vdash x : \tau}
\end{equation*}
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x \!:\! \tau_1}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x . e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2}
\end{equation*}
\begin{equation*}
\tyrule{app}
  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 \quad
    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2}
\end{equation*}
\begin{equation*}
\tyrule{let}
  { \coctx{\Gamma_1, x \!:\! \tau_1}{\aclrd{\textbf{r}} \times \langle{\cclrd{t}}\rangle} \vdash e_1 : \tau_2 \quad
    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \times (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}
\end{equation*}
\begin{equation*}
\tyrule{sub}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}
  {\coctx{\Gamma}{\aclrd{\textbf{r'}}} \vdash e : \tau}~\aclrd{\textbf{r}} \leq \aclrd{\textbf{r'}}
\end{equation*}
\vspace{0.5em}

@@

\paragraph{Summary.}
Comparing the structural coeffect calculus for tracking liveness and for bounded variable reuse
reveals which parts of the systems differ and which parts are shared. In particular, both systems
use the same vector operations ($\times$, $\alift{\cclrd{\textnormal{--}}}$) and also share the
lambda abstraction rule (\emph{abs}). They differ in the primitive values used to annotate used
and unused variables (\cclrd{L}, \cclrd{D} and $1$, $0$, respectively) and in the operators used
for sequential composition and contraction ($\cclrd{\sqcup}$, $\cclrd{\sqcap}$ and $\ast$, $+$, respectively).
The algebraic structure capturing these operators is developed in Chapter~\ref{ch:structural}.

>>>>

Effect systems generally use judgements of the form $\Gamma \vdash e : \tau \;\&\; \cclrd{\sigma}$,
associating effects $\sigma$ with the output type. We write coeffect
systems using judgements of the form $\coctx{\Gamma}{\cclrd{\sigma}} \vdash e : \tau$, associating
the context demands with $\Gamma$. Thus, we extend the traditional notion of free-variable
context $\Gamma$ with richer notions of context. This notation emphasizes the right intuition,
but there are more important differences between effects and coeffects.

@@

In Section~\ref{sec:intro-context-example}, we used \emph{resources} in a distributed system as an
example of flat coeffects. These could be, for example, a database, GPS sensor or access to the current
time. We also outlined that such context demands can be tracked as part of the typing assumption,
for example, say we have an expression $e$ that requires GPS coordinates and the current time.
The variable context of such expression will be annotated with a set of required resources,
\ie~$\coctx{\Gamma}{ \cclrd{\ident{\{ gps, time \}}} }$.

@@

We look at resources in distributed programming in more details in Section~\ref{sec:applications-flat-distr}.
The important point here is that in flat coeffect systems, contextual requirements are
\emph{split} between the call site and declaration site. Furthermore, there is no syntactic
structure that determines how the requirements are split. As mentioned in Section~\ref{sec:path-binding-amb},
we decouple the definition of semantics from the domain-specific choice that determines how
context demands are satisfied. We capture the choice in the type and give semantics over a
\emph{typing derivation}. A domain-specific algorithm then chooses the desirable typing -- for
example, by preferring resources available on the client over resources available on the server.

@@

Just as in the usual pointwise multiplication of a vector by a scalar, we lift any binary operation on scalars into a
scalar-vector one. For a binary operation on scalars $\circ : \C \times \C \rightarrow \C$, we define
 $\cclrd{s} \circ \aclrd{\textbf{r}} = \alift{ \cclrd{s}\circ\cclrd{r_1}, \ldots, \cclrd{s}\circ\cclrd{r_n}}$.
Relations on scalars can be also lifted to vectors. Given two vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}$ of the
same shape with positions $\{ 1, \ldots, n \}$ and a relation $\varpropto\, \subseteq \C \times \C$ we define
$\aclrd{\textbf{r}} \varpropto \aclrd{\textbf{s}} \Leftrightarrow (\cclrd{r_1} \varpropto \cclrd{s_1}) \wedge \ldots \wedge (\cclrd{r_n} \varpropto \cclrd{s_n}) $
Finally, we often concatenate vectors, for example, when joining two variable contexts.
Given vectors $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}$ with (possibly different) shapes $\{ 1, \ldots, n \}$ and
$\{ 1, \ldots, m \}$, the associative operation for concatenation $\vtimes$ is defined as
$\aclrd{\textbf{r}}\vtimes\aclrd{\textbf{s}} = \alift{\cclrd{r_1},\ldots,\cclrd{r_n},\cclrd{s_1},\ldots,\cclrd{s_m}}$.

We note that an environment $\Gamma$ containing $n$ uniquely named, typed variables is also a vector,
but we continue to write `$,$' for the product, so $\Gamma_1, x\!:\!\tau, \Gamma_2$ should
be seen as $\Gamma_1 \vtimes \langle x\!:\!\tau\rangle \vtimes \Gamma_2$.

@@

Accessing an ordinary variable (\emph{var}) does not require any implicit parameters. The rule that
introduces primitive context demands is (\emph{param}). Accessing a parameter \ident{?param}
requires it to be available in the context. The context may provide more (unused)
implicit parameters thanks to the subcoeffecting rule (\emph{sub}).

When we read the rules from the top to the bottom, application (\emph{app}) and let binding
(\emph{let}) simply union the context demands of the sub-expressions. However, lambda abstraction
(\emph{abs}) is where the example differs from effect systems. The implicit parameters required by
the body $\cclrd{r} \cup \cclrd{s}$ can be freely split between the declaration site ($\coctx{\Gamma}{\cclrd{r}}$)
and the call site ($\tau_1 \xrightarrow{\cclrd{s}} \tau_2$) and thus an expression may have multiple
valid typing derivations. Finally, (\emph{letdyn}) removes the bound parameter from the set of requirements.

@@

Type classes and implicit parameters show two important points about flat coeffect systems.
First, the context demands are associated with some \emph{scope}, such as the body
of a function. Second, they are associated with the input. To call a function that takes an
implicit parameter or has a type-class constraint, the caller needs to pass a (hidden) parameter
together with the function inputs.

@@

\paragraph{Summary.}
Unlike with implicit parameters, lambda abstraction for liveness analysis does not introduce
non-determinism. It simply duplicates the context demands. However, this still matches the
property of coeffects that impurities cannot be delayed or thunked and attached just to the
function arrow -- we place requirements on both call site and declaration site.

@@

\begin{figure}[t]
a.) Ordinary, syntax-driven rules along with subcoeffecting
\begin{equation*}
\tyrule{var}{}
  {\coctx{x \!:\! \tau}{\alift{\cclrd{1}}} \vdash x : \tau}
\end{equation*}
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x \!:\! \tau_1}{\aclrd{\textbf{r}} \cons \alift{\cclrd{s}}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash \lambda x . e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2}
\end{equation*}
\begin{equation*}
\tyrule{app}
  { \coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 \quad
    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \cons (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash e_1 \, e_2 : \tau_2}
\end{equation*}
\begin{equation*}
\tyrule{let}
  { \coctx{\Gamma_1, x \!:\! \tau_1}{\aclrd{\textbf{r}} \atimes \langle{\cclrd{t}}\rangle} \vdash e_1 : \tau_2 \quad
    \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash e_2 : \tau_1}
  {\coctx{\Gamma_1, \Gamma_2}{\aclrd{\textbf{r}} \atimes (\cclrd{t} \ast \aclrd{\textbf{s}})} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}
\end{equation*}
\begin{equation*}
\tyrule{sub}
  {\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau}
  {\coctx{\Gamma}{\aclrd{\textbf{r'}}} \vdash e : \tau}~\aclrd{\textbf{r}} \leq \aclrd{\textbf{r'}}
\end{equation*}
\vspace{0.5em}

@@

\paragraph{Summary.}
Comparing the structural coeffect calculus for tracking liveness and for bounded variable reuse
reveals which parts of the systems differ and which parts are shared. In particular, both systems
use the same vector operations ($\atimes$, $\alift{\cclrd{\textnormal{--}}}$) and also share the
lambda abstraction rule (\emph{abs}). They differ in the primitive values used to annotate used
and unused variables (\cclrd{L}, \cclrd{D} and $1$, $0$, respectively) and in the operators used
for sequential composition and contraction ($\cclrd{\sqcup}$, $\cclrd{\sqcap}$ and $\ast$, $+$, respectively).
The algebraic structure capturing these operators is developed in Chapter~\ref{ch:structural}.

### 2016-05-29 text/Chapter04.tex
Addressing feedback, fix demands (Chapter 6)

\subsection{Implicit parameters}

@@

The focus of this chapter is on the general coeffect system and so we do not discuss the
domain-specific reduction rules for individual context-aware language. Some work on operational
semantics of general coeffect systems has been done by Brunel et al. \cite{coeffects-quantitative}.
We give formal semantics of implicit parameters and dataflow in Chapter~\ref{ch:semantics} by
translation to a simple functional programming language instead.)

@@

\begin{lemma}[Bottom-pointed substitution]
\label{thm:cbn-substitution-bot}
In a bottom-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$
where $\czip = \cseq = \cpar$ is an idempotent and commutative operation`  ' and
$\cclrd{r}\,\cleq\,\cclrd{r'} \Rightarrow \forall\cclrd{s}.\cclrd{r}\,\cseq\,\cclrd{s}\;\cleq\;\cclrd{r'}\,\cseq\,\cclrd{s}$ then:
%
\begin{equation*}
\begin{array}{l}
\coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s \;\; \wedge \;\;
\coctx{\Gamma_1,  x : \tau_s, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau_r\\
\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} \,\cseq\, \cclrd{s} } \vdash \subst{e_r}{x}{e_s} : \tau_r
\end{array}
\end{equation*}

>>>>

\subsection{Implicit parameters}
\label{sec:flat-unique-impl}

@@

The focus of this chapter is on the general coeffect system and so we do not discuss the
domain-specific reduction rules for individual context-aware language. Some work on both operational
and denotational semantics of general coeffect systems has been done by Brunel et al.
\cite{coeffects-quantitative} and Breuvart and Pagani \cite{coeffects-modelling}.
We give formal semantics of implicit parameters and dataflow in Chapter~\ref{ch:semantics} by
translation to a simple functional programming language instead.)

@@

\begin{lemma}[Bottom-pointed substitution]
\label{thm:cbn-substitution-bot}
In a bottom-pointed flat coeffect calculus with an algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$
where $\czip = \cseq = \cpar$ is an idempotent and commutative operation and
$\cclrd{r}\,\cleq\,\cclrd{r'} \Rightarrow \forall\cclrd{s}.\cclrd{r}\,\cseq\,\cclrd{s}\;\cleq\;\cclrd{r'}\,\cseq\,\cclrd{s}$ then:
%
\begin{equation*}
\begin{array}{l}
\coctx{\Gamma}{\cclrd{s}} \vdash e_s : \tau_s \;\; \wedge \;\;
\coctx{\Gamma_1,  x : \tau_s, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau_r\\
\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} \,\cseq\, \cclrd{s} } \vdash \subst{e_r}{x}{e_s} : \tau_r
\end{array}
\end{equation*}

### 2016-05-29 text/Chapter05.tex
Addressing feedback, fix demands (Chapter 6)

\item For two sample coeffect calculi discussed earlier (dataflow and implicit parameters),
  we give reduction rules for the comonadically-inspired primitives and we extend the progress
  and preservation proofs, showing that well-typed programs produced by translation from two
  coeffect languages do not get stuck (Section~\ref{sec:semantics-proofs})

>>>>

\item For two sample coeffect calculi discussed earlier (dataflow and implicit parameters),
  we give reduction rules for the comonadically-inspired primitives and we extend the progress
  and preservation proofs, showing that well-typed programs produced by translation from two
  coeffect languages do not go wrong (Section~\ref{sec:semantics-proofs})
